---
title: "Class 12 HW"
author: "Sarah Mirsaidi Madjdabadi, A16890186"
format: pdf
---

## Section 4: Population Scale Analysis
One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3 expression.

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
# total number of samples
```

```{r}
table(expr$geno)
# sample size for each genotype
```

A/A Median Expression level: 31.25
```{r}
summary(expr$exp[expr$geno == "A/A"])
```

A/G Median Expression level: 25.065

```{r}
summary(expr$exp[expr$geno == "A/G"])
```

G/G Median Expression level: 20.074

```{r}
summary(expr$exp[expr$geno == "G/G"])
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)
```

```{r}
ggplot(expr) + aes(x=geno, y=exp, fill=geno) +
  geom_boxplot(notch=TRUE) +
  labs(x = "Genotype", y = "Expression Level")
```
Having G/G in this location is associated with having a reduced expression of ORMDL3 (no overlap in notches, significant difference); yes, SNP does impact gene expression.
